Numerical simulation of the Cordilleran ice sheet through the last glacial cycle

### 2014-05-23 cordillera-cycle.tex
Corrections from Irina in abstract and intro.

% My commands
\def\idea#1{\textcolor[rgb]{0,0.5,0}{\textbf{[IDEA: #1]}}}
\def\todo#1{\textcolor[rgb]{0.5,0,0}{\textbf{[TODO: #1]}}}
\def\aref{\textcolor[rgb]{0.5,0,0}{\textbf{[REF]}}}

@@

  Despite more than a century of geological observations, the Cordilleran ice
  sheet of North America remains poorly understood in terms of its former
  extent, volume and dynamics. Although geomorphological evidence is abundant,
  its complexity is such that whole ice-sheet reconstructions of advance and
  retreat patterns are lacking. Here we use a numerical ice sheet model
  calibrated against field-based evidence to attempt a quantitative
  reconstruction of the Cordilleran ice sheet history through the last glacial
  cycle. A series of simulations is driven by time-dependent temperature
  offsets from six proxy records located around the globe. Although this
  approach reveals large variations in model response to evolving atmospheric
  forcing, all simulations produce two major glaciations events during MIS~4
  (61.9--55.4\,\unit{kyr}) and MIS~2 (29.5--16.9\,\unit{kyr}). The timing of
  glaciation is
  better reproduced using temperature forcing derived from Greenland and
  Antarctic ice-core than from regional ocean sediment cores. During most of
  the last glacial cycle, the modelled ice cover is discontinuous and
  restricted to high mountain areas. However, widespread precipitation over the
  Skeena mountains favours the persistence of a central ice dome throughout the
  glacial cycle. It acts as a nucleation centre before the last glacial maximum
  and hosts the last remain of Cordilleran ice during the
  early Holocene (10.9--9.5\,\unit{kyr}).

@@

During the last glacial cycle, glaciers and ice caps of the North American
Cordillera have been more extensive than today. At the peak of glaciation, a
contiguous blanket of ice extended from Denali Peak in the north to the Puget
Sound in the south. This blanket is known as the former Cordilleran ice sheet.

For more than a century, exploration and geological investigation of the
Cordillera lead to a large collection of evidence of the former ice cover. This
evidence consists of mapped boundaries of the former ice extent, direction of
past sliding velocities and location of former melt-water streams, as well as
carbon dating and cosmogenic dating of glacier retreat.

Field-based evidence allowed to reconstruct, with a high degree of confidence,
the maximal extent attained by the Cordilleran ice sheet during the last
glacial cycle. However, former ice thickness, and in turn the ice sheet's
contribution to the last glacial maximum sea-level low-stand, remains poorly
constrained. Moreover, our understanding of the deglaciation phase is
restricted to a regional level, while little is known about the ice sheet
history preceding the last glacial maximum extent. Although time-evolving,
whole ice-sheet reconstructions of glacial advance and retreat patters are
available for the Laurentide and Eurasian ice sheet, this is not the case for
the Cordilleran ice sheet.

The present study aims to use a numerical ice sheet model calibrated against
field-based evidence to attempt a quantitative reconstruction of the
Cordilleran ice sheet history through the last glacial cycle. Although
numerical modelling has been identified as a tool to improve our understanding
of Cordilleran ice sheet more than twenty years ago, the ubiquitously
mountainous topography of the region has presented a major challenge to its
application, which is only overcome by recent development of numerical ice
sheet models and underlying scientific tools. In addition to the topographic
complexity, strong climatic contrasts characteristic of the North American
Cordillera require the use of high-resolution temperature and precipitation
forcing as input data to the ice sheet model.

Here we use the Parallel Ice Sheet Model (PISM), forced by a high-resolution
climatology from the North American Regional Reanalysis, to model the
Cordilleran ice sheet through the last glacial cycle. Because past climate
is mostly unknown, palaeoclimate forcing is applied in the simplified form of
six time-dependent temperature offset time-series. These time-series were
obtained by scaling six different palaeo-temperature reconstructions from
proxy records around the globe, including two \chem{\delta^{18}O} records from
Greenland ice cores, two \chem{\delta^{18}O} records from Antarctic ice cores,
and two alkenone unsaturation index records from Northwest Pacific oceanic
sediment cores. The output from the model is compared to geomorphological
evidence in terms of timing of glaciation, extent of glaciation and patterns
of deglaciation.

>>>>

% My commands
\def\idea#1{\textcolor[rgb]{0,0.5,0}{\textbf{[IDEA: #1]}}}
\def\todo#1{\textcolor[rgb]{0.5,0,0}{\textbf{[TODO: #1]}}}
\def\aref{\textcolor[rgb]{0.5,0,0}{\textbf{[REF]}}}
\def\citep#1{}
\def\citet#1{}

@@

  Despite more than a century of geological observations, the Cordilleran ice
  sheet of North America remains poorly understood in terms of its former
  extent, volume and dynamics. Although geomorphological evidence is abundant,
  its complexity is such that whole ice-sheet reconstructions of advance and
  retreat patterns are lacking. Here we use a numerical ice sheet model
  calibrated against field-based evidence to attempt a quantitative
  reconstruction of the Cordilleran ice sheet history through the last glacial
  cycle. A series of simulations is driven by time-dependent temperature
  offsets from six proxy records located around the globe. Although this
  approach reveals large variations in model response to evolving atmospheric
  forcing, all simulations produce two major glaciation events during MIS~4
  (61.9--55.4\,\unit{kyr}) and MIS~2 (29.5--16.9\,\unit{kyr}). The timing of
  glaciation is
  better reproduced using temperature reconstructions from Greenland and
  Antarctic ice cores than from regional ocean sediment cores. During most of
  the last glacial cycle, the modelled ice cover is discontinuous and
  restricted to high mountain areas. However, widespread precipitation over the
  Skeena mountains favours the persistence of a central ice dome throughout the
  glacial cycle. It acts as a nucleation centre before the last glacial maximum
  and hosts the last remains of Cordilleran ice during the
  early Holocene (10.9--9.5\,\unit{kyr}).

@@

During the last glacial cycle, glaciers and ice caps of the North American
Cordillera were more extensive than today. During the last glacial maximum, a
contiguous blanket of ice extended from Denali Peak in the north to the Puget
Sound in the south. This ice mass is known as the former Cordilleran ice
sheet (Fig.~\ref{fig:locmap}).

For more than a century, exploration and geological investigation of the
Cordillera have led to a large collection of evidence of the former ice cover. This
evidence consists of mapped boundaries of the former ice extent, directions of
past sliding velocities and locations of former melt-water streams, as well as
the timing of glacier retreat identified from radiocarbon and cosmogenic dating.

Field-based evidence allowed to reconstruct, with a high degree of confidence,
the maximal extent attained by the Cordilleran ice sheet during the last glacial
cycle. However, former ice thickness and the ice sheet's contribution to the
last glacial maximum sea-level low-stand remain poorly constrained. Moreover,
our understanding of the glaciation history is mainly restricted to the
deglaciation phase and a regional level, while little is known about the ice
sheet evolution prior to the last glacial maximum extent. Although
time-evolving, whole ice-sheet reconstructions of glacial advance and retreat
patters are available for the Laurentide and Eurasian ice sheets, such is not
the case for the Cordilleran ice sheet, where complex patterns of ice flow
emanating from multiple glaciation centres have been identified.

The present study uses a numerical ice sheet model (PISM) calibrated against
field-based evidence to perform a quantitative reconstruction of the
Cordilleran ice sheet history through the last glacial cycle. Although
numerical modelling has been established as a useful tool to improve our
understanding
of the Cordilleran ice sheet more than twenty years ago \aref, the ubiquitously
mountainous topography of the region has presented a major challenge to its
application, which has only been overcome by recent development of numerical ice
sheet models and underlying scientific computing tools. In addition to the topographic
complexity, strong climatic contrasts characteristic of the North American
Cordillera require the use of a high-resolution temperature and precipitation
forcing as input data to the ice sheet model.

Because past climate conditions are subject to considerable uncertainty, our
palaeoclimate forcing is a simplistic approximation of time-evolved temperature
and precipitation fields derived from a combination of a present-day
atmospheric reanalysis (NARR), lapse-rate corrections and temperature offset
time series. The latter are
obtained by scaling six different palaeo-temperature reconstructions from
proxy records around the globe, including two \chem{\delta^{18}O} records from
Greenland ice cores, two \chem{\delta^{18}O} records from Antarctic ice cores,
and two alkenone unsaturation index records from Northwest Pacific oceanic
sediment cores. Model output is compared to geomorphological evidence in terms
of timing and extent of glaciation and patterns of deglaciation.
